Tool-driven Revolutions in Archaeological Science

### 2018-09-09 analysis/paper/jcaa-paper.Rmd
lots more text

---
title: "Tool-driven Revolutions in Archaeological Science"
author:
  - Sophie C. Schmidt
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
abstract: "There is an argument in philosophy of science that revolutions in science are either ideas-driven or tool-driven. We explore this debate in light of recent efforts by many scientific disciplines to embrace methods to improve the reproducibility of their research. One of the most profound changes driven by this concern for reproducibility and transparency is from analysing data using tools dependent on pointing-and-clicking with a mouse in commercial software, to a new set of tools based on writing scripts in free and open source programming languages. We review bibliometric evidence for this change in the sciences generally and in archaeology in particular to test if the adoption of these new tools may be considered a revolution or some other process. We evaluate how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches."
---

@@

In archaeology we have seen claims for paradigm shifts in several areas. The most wide-ranging application is Clark's [-@Clark_1993] comparison of North American archaeology to European archaeology during the 1970s-90s. Clark, similar to Binford and Sabloff [-@binford1982paradigms], identified the North American archaeological paradigm as rooted in anthropology, founded on post-positivist biases, and distinctive with it critical-realist ontology, a modified-objectivist epistemology, and an experimental-manipulative methodology. In contrast, the European archaeological paradigm is described a kind of history, mostly in a strict empiricist tradition, and typified by a realist ontology, an inductivist epistemology, and by an observational methodology. Clark's analysis describes a multi-paradigm situation where multiple research communities (i.e. North American and European) simutaneously operate in different paradigms (which Kuhn noted as a sign of a immature discipline). His account is not a complete Kuhnian analysis because it lacks a revolutionary change event where one paradigm is replaced by another, although he hints that processual archaeology may be considered a radical discontinuity relative to the culture-historical approaches that preceeded it [@Clark_1993, pp. 206].

@@

Galison's emphasis on tool-driven change, as a complement to Kuhnian concept-driven change, has much potential for archaeology. The centrality of material culture in Galison's view raises the question about whether there is scope for archaeologists to make productive contributions to understanding change in the history science through an archaeological analysis of scientific instrumentation [cf. @Schiffer_2013]. Furthermore, Galison's analysis of the image/logic contrast in the history of physics invites a similar analysis of the history of archaeology Can we identify archaeological traditions focused on data collection and analysis using images and image-making devices [cf. @molyneaux2013cultural], in contrast to digital devices? One candidate for this might be the shift from building relative chronologies based on seriation of visually distinctive artefacts to absolute chronologies based on radiometric dating using computer-controlled instruments. A third issue is how we can use a tool-driven approach not just as a framework for understanding change in science, but as a method to generate change in archaeology?

# Tools Emerging from the Reproducibility Crisis

We explore this debate in light of recent efforts by many scientific disciplines to embrace methods to improve the reproducibility of their research. One of the most profound changes driven by this concern for reproducibility and transparency is from analysing data using tools dependent on pointing-and-clicking with a mouse in commercial software, to a new set of tools based on writing scripts in free and open source programming languages...

# Tool-adoption in archaeology and elsewhere

We review bibliometric evidence for this change in the sciences generally and in archaeology in particular to test if the adoption of these new tools may be considered a revolution or some other process. We evaluate how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches.

rrtools... huskydown

special role of archaeology here, since we are concerned about past behaviours, we have a special interest in research leaving  a visible trace of our work

## Slide: Articles in all sciences citing R

```{r all-sciences-citing-r}

# We did a 'Cited Reference Search' in WoS for title "R: A Language and Environment for Statistical Computing" and author = "R Core Team", which is the output from  `citation()`. This function first appeared in 2004 https://github.com/wch/r-source/commits/9a5c086855d2e8fb07f670e3315871b2de31a4fd/src/library/base/inst/CITATION 
# 
# Note that before 2012 the authors were "R Development Core Team", and then it was simplified to "R Core Team" https://github.com/wch/r-source/commit/9a5c086855d2e8fb07f670e3315871b2de31a4fd#diff-80bb8a8b133a6204e9642d76791927e2
# 
# This is our exact search string for WoS:
# 
# You searched for: CITED AUTHOR: ("R DEV COR TEAM" OR "R CORE" OR "R CORE TEAM") AND CITED WORK: ("R: A Language and Environment for Statistical Computing") AND LANGUAGE: (English) AND DOCUMENT TYPES: (Article)
# Timespan: All years. Indexes: SCI-EXPANDED, SSCI, A&HCI, ESCI.

cran_cites_files <- 
    paste0(here::here(), glue("/analysis/data/raw_data/cited_cran/savedrecs ({1:11}).txt"))

@@

min_y <- 2007
max_y <- 2025
all_sciences_citing_R_over_time <- 
ggplot(cran_cites_all_areas_top_journals_by_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_text_repel(
    data = subset(cran_cites_all_areas_top_journals_by_year, 
                  PY == max(PY)),
    aes(label = str_wrap(SO, 40)),
    size = 4,
    nudge_x = 0,
    hjust = 0,
    segment.color = NA,
    direction = "y"
  ) +
   scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, rep("", length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  ggtitle(str_glue('Percentage of articles per year citing R in top 10 WoS journals ({prettyNum(how_many_articles_in_top_journals_for_cran_cites,",")} articles)'),
          subtitle = "Data from apps.webofknowledge.com,\nusing in SCI-EXPANDED & SSCI for 2007-2017")

@@

This plot shows how data analysis is changing in the sciences more broadly towards the use of the R programming language. We searched the Web of Science, produced by Clarivate Analytics, for citations to the R program over the last ten years, and filtered the results to the top ten journals with the highest number of citations to this software. The purpose of this filtering step was to find areas of research were R is generally known, so we can see how its use is changing over time. Note that many articles that use R do not cite the R program, so these numbers underestimate the true use of R in science.

We can make two major observations:

**First**, the areas of science that have seen big growth in the use of R are ecology, evolution and related biological sciences. This is especially significant for us archaeologists, as ecology and biology are subjects from which we quite often adapt methods for analysis, as for example in point-pattern-analysis and statistical modelling. 
We also see increases in multi-disciplinary journals such as PeerJ and PLOS One.

**Second**, the scale of the increases is quite impressive: Starting only 2012 some journals now have 20% of articles citing R.

So, we can see this is a substantial change in the way researchers are working in many fields. What about archaeology? Do we also see a change like this?

## Slide: Archaeology articles citing R

@@

# get the top archaeology journals by numbers of articles.
wc_archaeology_2017 <- 
  readr::read_tsv(paste0(here::here(), "/analysis/data/raw_data/savedrecs.txt"), 
            quote = "")

# The 10 journals with the most articles are:
top_journals_by_articles <- 
wc_archaeology_2017 %>% 
  group_by(SO) %>% 
  tally(sort = TRUE)

# For 2007-2017:

data_files_from_2007_to_2017 <- 
  paste0(here::here(), glue("/analysis/data/raw_data/savedrecs ({1:2}).txt"))

wc_archaeology_2007_to_2017 <- 
  map(data_files_from_2007_to_2017,
      ~readr::read_tsv(.x, quote = "")) %>% 
  bind_rows()

total_number_of_articles_in_sample <- 
  nrow(wc_archaeology_2007_to_2017)

# most populararchaeology journals over time, that have some mention of R
most_popular_journals_over_time <- 
wc_archaeology_2007_to_2017 %>% 
  group_by(SO, PY) %>% 
  tally(sort = TRUE) %>% 
  filter(!SO %in% c("JOURNAL OF CULTURAL HERITAGE",
                    "STUDIES IN CONSERVATION",
                    "INTERSECCIONES EN ANTROPOLOGIA",
                    "ISRAEL EXPLORATION JOURNAL",
                    "JOURNAL OF MATERIAL CULTURE",
                    "TRABAJOS DE PREHISTORIA",
                    "ARCHEOSCIENCES-REVUE D ARCHEOMETRIE")) %>% 
  filter(n >= 2) %>% 
  ungroup()

most_popular_journals_over_time_names <- 
  unique(most_popular_journals_over_time$SO)

archy_journals_total_number_of_articles <- 
  read_csv(str_glue('{here::here()}/analysis/data/raw_data/top_archaeology_journals_articles_per_year.csv'))

top_archaeology_journals_articles_per_year <- 
  archy_journals_total_number_of_articles %>% 
  gather(variable, value, -journal_name) %>% 
  mutate(year = as.numeric(variable)) %>% 
  select(-variable)

cran_cites_in_archy_journals <- 
  cran_cites_all_areas_all_years %>% 
  filter(SO %in% most_popular_journals_over_time_names)

# what archaeology journals cite R?
# unique(cran_cites_in_archy_journals$SO)

archy_journals_total_number_of_citing_articles <- 
  nrow(cran_cites_in_archy_journals)

cran_cites_in_archy_journals_per_year <- 
  cran_cites_in_archy_journals %>% 
  group_by(SO, PY) %>% 
  tally(sort = TRUE) %>% 
  ungroup() %>% 
  mutate(PY = as.numeric(PY))  %>% 
  left_join(top_archaeology_journals_articles_per_year,
            by = c("SO" = "journal_name",
                   "PY" = "year")) %>% 
  mutate(prop = n / value) %>% 
  group_by(SO) %>% 
  mutate(jn = str_glue('{SO} (n = {sum(n)})')) %>% 
  ungroup()

@@

jas_cites_r <- 
cran_cites_in_archy_journals_per_year %>% 
  filter(SO == "JOURNAL OF ARCHAEOLOGICAL SCIENCE") %>% 
  ggplot(aes(PY,
           prop)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm",
              se = FALSE) +
  stat_poly_eq(aes(label =  paste(..eq.label.., 
                                  ..adj.rr.label..,
                                  sep = "~~~~")),
               formula = y~x, 
               parse = TRUE,
               size = 3,
               label.y.npc = 0.7) +
  stat_fit_glance(label.y.npc = 0.7,
                  size = 3,
                  method = "lm", 
                  method.args = list(formula = y ~ x),
                  geom = "text",
                  aes(label = paste("p-value: ",signif(..p.value.., digits = 4)))) +
  xlab("year") +
  ylab("Proportion of articles citing R") +
  ggtitle(expression(paste("Articles citing R in ", italic("Journal of Archaeological Science")))) +
  theme_bw(base_size = 6)

all_archaeology_citing_R_over_time <- 
ggplot(cran_cites_in_archy_journals_per_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_text_repel(
    data = subset(cran_cites_in_archy_journals_per_year, 
                  PY == max(PY)),
    aes(label = jn),
    direction = "y",
    size = 5,
    nudge_x = 1,
    hjust = 0) +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, rep("", length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal") +
  ggtitle(str_glue('Percentage of Archaeology articles per year citing R ({archy_journals_total_number_of_citing_articles} out of {prettyNum(total_number_of_articles_in_sample, ",")} articles in sample)'),
          subtitle = "Data from apps.webofknowledge.com,\nusing in SCI-EXPANDED & SSCI for 2007-2017")  +
  annotation_custom(grob = ggplotGrob(jas_cites_r), 
                          xmin = 2030, xmax = 2040, 
                          ymin = 0.10, ymax = 0.17)

ggsave(plot = all_archaeology_citing_R_over_time, 
       filename = "all_archaeology_citing_R_over_time.png",
       path = str_glue('{here::here()}/analysis/figures'),
       height = 6,
       width = 10)

knitr::include_graphics(str_glue('{here::here()}/analysis/figures/all_archaeology_citing_R_over_time.png'))
```

Here are the results for archaeological articles. We found `r archy_journals_total_number_of_citing_articles` articles that cite R over the last ten years. Note that this is certainly not complete, since these are just the journals tracked by the Web of Science database, and we know that archaeologists publish in a great variety of journals.

There are three important details in these plots:

The **first** detail we see here is that overall, the proportions and the absolute number of articles citing R in archaeology are much smaller than what we see in the ecology journals on the previous slide. Only the _Journal of Archaeological Science_ has more than 10 articles in our sample. It's obvious, we have yet to adopt programming for data analysis in the same way that ecologists have.

The **second** detail is that we do not see a significant increase over time. Both archaeology and the other sciences we saw on the previous slide show an increase in citing R at around 2013-2014, but we do not yet see much of an upward trend in the archaeology journals. Indeed, in several of our journals the first citations of R only occured in the last two years. The plot on the upper right of this slide shows a statistical test of change over time for articles in JAS. With a low r-squared value and high p-value for the linear model, we conclude that there is no evidence for an increase in citations over time in this journal.

The **third** detail we see is the distinctive types of journals that have articles citing R. There is a clear focus on scientific archaeology, with _Journal of Archaeological Science_ having the highest total number of articles. Archaeology as a subject has always been situated between a humanistic and a scientific tradition, but studies using statistical methods lean towards the sciences, which we mean with the term "scientific archaeology". We also see geoarchaeology, lithics, and environmental archaeology represented in the journal names. We can investigate this further with statistical analysis of the words in the titles of all the articles in our sample.

## Slide: What are archaeologists doing with R?

```{r text-mining-article-titles-archaeology}
# https://www.tidytextmining.com/twitter.html

@@

wc_archaeology_2007_to_2017_no_R_cites <- 
  wc_archaeology_2007_to_2017 %>% 
  filter(TI %nin% cran_cites_in_archy_journals$TI)

# combine
archy_journal_articles <- 
  bind_rows(wc_archaeology_2007_to_2017_no_R_cites %>% 
            mutate(ID = "no_R_cites",
                   VL = as.character(VL)),
           cran_cites_in_archy_journals %>% 
            mutate(ID = "has_R_cites"))

tidy_cites <- 
archy_journal_articles %>% 
  unnest_tokens(word, TI) %>%
  filter(!word %in% c(stop_words$word, "archaeology", 
                      "archaeological",
                      "analysis", "study", "site"),
         str_detect(word, "[a-z]"))

@@

Here we see a comparison of word frequencies of the titles of archaeology journal articles. Words near the red line are used with about equal frequencies by papers that cite R and by papers that do not cite R. Words far away from the red line are used much more by one group of articles compared to the other.

>>>>

---
title: "Tool-driven Revolutions in Archaeological Science"
author:
  - Sophie C. Schmidt
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
abstract: "There is an argument in philosophy of science that revolutions in science are either ideas-driven or tool-driven. We explore this debate in light of recent efforts by many scientific disciplines to embrace methods to improve the reproducibility of their research. One of the most profound changes driven by this concern for reproducibility and transparency is from analysing data using tools dependent on pointing-and-clicking with a mouse in commercial software, to a new set of tools based on writing scripts in free and open source programming languages. We review bibliometric evidence for this change in the sciences generally and in archaeology in particular to test if the adoption of these new tools may be considered a revolution or some other process. We evaluate how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches."
---

@@

In archaeology we have seen claims for paradigm shifts in several areas. The most wide-ranging application is Clark's [-@Clark_1993] comparison of North American archaeology to European archaeology during the 1970s-90s. Clark, similar to Binford and Sabloff [-@binford1982paradigms], identified the North American archaeological paradigm as rooted in anthropology, founded on post-positivist biases, and distinctive with it critical-realist ontology, a modified-objectivist epistemology, and an experimental-manipulative methodology. In contrast, the European archaeological paradigm is described a kind of history, mostly in a strict empiricist tradition, and typified by a realist ontology, an inductivist epistemology, and by an observational methodology. Clark's analysis describes a multi-paradigm situation where multiple research communities (i.e. North American and European) simutaneously operate in different paradigms (which Kuhn noted as a sign of a developing or immature discipline). His account is not a complete Kuhnian analysis because it lacks a revolutionary change event where one paradigm is replaced by another, although he hints that processual archaeology may be considered a radical discontinuity relative to the culture-historical approaches that preceeded it [@Clark_1993, pp. 206].

@@

Galison's emphasis on tool-driven change, as a complement to Kuhnian concept-driven change, has much potential for archaeology. The centrality of material culture in Galison's view raises the question about whether there is scope for archaeologists to make productive contributions to understanding change in the history science through an archaeological analysis of scientific instrumentation [cf. @Schiffer_2013]. Furthermore, Galison's analysis of the image/logic contrast in the history of physics invites a similar analysis of the history of archaeology Can we identify archaeological traditions focused on data collection and analysis using images and image-making devices [cf. @molyneaux2013cultural], in contrast to digital devices? One candidate for this might be the shift from building relative chronologies based on seriation of visually distinctive artefacts to absolute chronologies based on radiometric dating using computer-controlled instruments. A third issue is how we can use a tool-driven approach not just as a framework for understanding change in science, but as a method to predict or generate change in archaeology?

# Bibliometric analysis of tool-driven change

## Looking back on a tool-driven change in Ecology

The question of predicting or generating change in scientific practice via adoption of new tools is especially intruiging, and in this section we explore how a change in ecology might be relevant for understanding the future of archaeology. @Touchon_McCoy_2016 investigated evaluated changes in statistics used by ecologists as a potential area of tool-driven change. They searched nearly 20,000 published articles in ecology published between 1990 and 2013. They found that there has been a rise in complex and computationally intensive statistical techniques such as mixed effects models and Bayesian statistics and a decline in reliance on approaches such as ANOVA or t-tests. Crucially, they found that ecologists have shifted away from software tools such as SAS and SPSS to the open source program R.

@Touchon_McCoy_2016 identify four factors relating to technological change that might explain the changes they observed in the use of statistics in ecology. First, they note that automated data loggers, GPS trackers, remote sensing, and crowd sourcing have greatly increased rate at which ecologists collect data. Second, increases in desktop and cluster computing power has made complex analytical processes faster and more convienent to compute. Third, the development of free, open source and easily extensible software for data analysis and visualisation, such as R, allows new methods to spread quickly via online fora and social media. Fourth, the publication of several books and papers that have strongly influenced the way many ecologists think about data analysis, such as @burnham2003model's book on model selection and inference.

While all of these four factors could be considered relevant to the tooling of ecology and archaeology, we focus here on the third one because of its relevance to archaeology. Archaeology and ecology each have a high diversity of data collection methods, and types of data analyses they conduct, with only a small overlap. What they share in common is working with the field-collected data on a computer to prepare it for publication. To investigate changes in the use of R in archaeology and ecology, we obtained reference lists from a sample of scholarly articles in the Web of Science database and examined patterns in the citation of the R program over time.

```{r ecology-citing-r}

# We did a 'Cited Reference Search' in WoS for title "R: A Language and Environment for Statistical Computing" and author = "R Core Team", which is the output from  `citation()`. This function first appeared in 2004 https://github.com/wch/r-source/commits/9a5c086855d2e8fb07f670e3315871b2de31a4fd/src/library/base/inst/CITATION 
# 
# Note that before 2012 the authors were "R Development Core Team", and then it was simplified to "R Core Team" https://github.com/wch/r-source/commit/9a5c086855d2e8fb07f670e3315871b2de31a4fd#diff-80bb8a8b133a6204e9642d76791927e2
# 
# This is our exact search string for WoS:

# Results: 38,040
# (from Web of Science Core Collection)
# You searched for: CITED AUTHOR: ("R DEV COR TEAM" OR "R # CORE" OR "R CORE TEAM" OR "R DEVELOPMENT CORE TEAM")
# Refined by: DOCUMENT TYPES: ( ARTICLE ) AND LANGUAGES: ( ENGLISH ) AND RESEARCH AREAS: ( ENVIRONMENTAL SCIENCES ECOLOGY )
# Timespan: 1900-2018. Indexes: SCI-EXPANDED, SSCI, A&HCI, # ESCI.

# field tags
# TS= Topic 
# TI= Title 
# AU= Author [Index] 
# AI= Author Identifiers 
# GP= Group Author [Index] 
# ED= Editor 
# SO= Publication Name [Index] 
# DO= DOI 
# PY= Year Published 
# AD= Address 
# OG= Organization-Enhanced [Index] 
# OO= Organization 
# SG= Suborganization 
# SA= Street Address 
# CI= City 
# PS= Province/State 
# CU= Country/Region 
# ZP= Zip/Postal Code 
# FO= Funding Agency 
# FG= Grant Number 
# FT= Funding Text 
# SU= Research Area 
# WC= Web of Science Category 
# IS= ISSN/ISBN 
# UT= Accession Number 
# PMID= PubMed ID

cran_cites_files <- 
    list.files(here::here("/analysis/data/raw_data/cited_cran/"),
               full.names = TRUE)

@@

articles_shown_in_plot <- 
  cran_cites_all_areas_top_journals_by_year %>% 
  filter(!is.na(prop)) %>% 
  summarise(sum_n = sum(n)) %>% 
  pull(sum_n)

# nice labels with n for plot
cran_cites_all_areas_top_journals_by_year <- 
cran_cites_all_areas_top_journals_by_year %>% 
  group_by(SO) %>% 
  summarise(total_n_articles = sum(n)) %>% 
  mutate(SO_n = as.character(glue('{SO} (n = {total_n_articles})'))) %>% 
  right_join(cran_cites_all_areas_top_journals_by_year,
             by = "SO")
```

We used the 'Cited Reference Search' function provided by the Web of Science online scientific citation indexing service to identify journal articles that cite R. Although R has been available since the late 1990s [@Thieme_2018], a recommended format for citing the software did not appear until 2004, with the author given as "R Development Core Team". This recommended format for citing the software changed slightly in 2012 when the author was updated to "R Core Team". We searched the Web of Science using '"R DEV COR TEAM" OR "R CORE" OR "R CORE TEAM" OR "R DEVELOPMENT CORE TEAM"' in the CITED AUTHOR field. We sorted the results by frequency of citation, and selected the first 1000 items (the maximum allowed by the Web of Science service). These 1000 items represent variations on the recommended format for citing R, and found in the reference lists of `r prettyNum(how_many_articles_in_top_journals_for_cran_cites, ",")` English-languge articles indexed by the Web of Science in the research area of Environmental Sciences Ecology. We then downloaded the bibliographic data for all of these articles.

Figure \@ref(fig:ecology-citing-r-plot) shows the percentage of articles citing R in each of several of the top ecology journals (as defined by how often their articles are cited). We restrict the start of the observation period to 2007 for convivnence. The plot shows that the percentage of articles citing R has increased from less than ten percent in all journals in the late 2000s, to more than 30% in _Ecosphere_, _Ecology and Evolution_, and _Molecular Ecology_ after 2012. We would not call this a Kuhnian paradigm shift, but it does show a substantial change in the tools of the discipline, supporting claims for a Galisonian tool-driven change in ecology similar to the changes described by @Touchon_McCoy_2016.

```{r ecology-articles-citing-r-plot, fig.cap=str_glue('Percentage of articles per year citing R in top Ecology journals ({prettyNum(articles_shown_in_plot,",")} articles out of {prettyNum(how_many_articles_in_top_journals_for_cran_cites,",")}). Data from Web of Science for 2007-2017')}
min_y <- 2007
max_y <- 2025
all_sciences_citing_R_over_time <- 
ggplot(cran_cites_all_areas_top_journals_by_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_text_repel(
    data = subset(cran_cites_all_areas_top_journals_by_year, 
                  PY == max(PY)),
    aes(label = str_wrap(SO_n, 40)),
    size = 4,
    nudge_x = 0,
    hjust = 0,
    segment.color = NA,
    direction = "y"
  ) +
   scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, rep("", length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal in that year") +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

@@

## Looking forward to a tool-driven change in Archaeology

@@

# Results: 44,047
# (from Web of Science Core Collection)
# You searched for: (WC=Archaeology) AND LANGUAGE: (English) AND DOCUMENT TYPES: # (Article)
# Timespan: 1900-2018. Indexes: SCI-EXPANDED, SSCI, A&HCI, ESCI

#- get journals with the most number of articles

archy_files <- 
    list.files(here::here("/analysis/data/raw_data/wos_archaeology/"),
               full.names = TRUE)

archy_all_years <- 
  map(archy_files,
      ~readr::read_tsv(.x, 
                       quote = "",
                       col_types = cols(.default = col_character()))) %>% 
  bind_rows() %>% 
  mutate(PY = as.numeric(PY)) %>% 
  filter(PY <= 2017)

# limit to top journals
top_journals_for_archy<- 
  archy_all_years %>% 
  group_by(SO) %>% 
  tally(sort = TRUE)

# n articles in top journals per year
top_journals_for_archy_per_year <- 
  archy_all_years %>% 
  group_by(SO, PY) %>% 
  tally() %>% 
  arrange(PY)

how_many_articles_in_top_archy_journals <- 
  sum(top_journals_for_archy$n)

#-- get articles citing R that are archaeology

# Results: 137
# (from Web of Science Core Collection)
# You searched for: CITED AUTHOR: ("R DEV COR TEAM" OR "R CORE" OR "R CORE TEAM" OR "R DEVELOPMENT CORE TEAM")
# Refined by: WEB OF SCIENCE CATEGORIES: ( ANTHROPOLOGY ) AND WEB OF SCIENCE CATEGORIES: ( ARCHAEOLOGY )
# Timespan: All years. Indexes: SCI-EXPANDED, SSCI, A&HCI, ESCI

archy_cran_files <- 
      list.files(here::here("/analysis/data/raw_data/cited_cran_archaeology/"),
               full.names = TRUE)

archy_cran_all_years <- 
  map(archy_cran_files,
      ~readr::read_tsv(.x, 
                       quote = "",
                       col_types = cols(.default = col_character()))) %>% 
  bind_rows() %>% 
  mutate(PY = as.numeric(PY)) %>% 
  filter(PY <= 2017)

# tally by year
archy_cran_cites_all_areas_top_journals_by_year <- 
  archy_cran_all_years %>% 
  filter(SO %in% top_journals_for_archy$SO) %>% 
  group_by(SO, PY) %>% 
  tally(sort = TRUE)  %>% 
  left_join(top_journals_for_archy_per_year,
            by = c('SO', 'PY')) %>% 
   mutate(prop = n.x / n.y) %>% 
  ungroup()

total_number_archaeology_articles_citing_r <- sum(archy_cran_cites_all_areas_top_journals_by_year$n.x)

# check for slight differences in journal names
# sort(unique(archy_cran_cites_all_areas_top_journals_by_year$SO))

# nice labels with n for plot
archy_cran_cites_all_areas_top_journals_by_year <- 
archy_cran_cites_all_areas_top_journals_by_year %>% 
  group_by(SO) %>% 
  summarise(total_n_articles = sum(n.x)) %>% 
  mutate(SO_n = as.character(glue('{SO} (n = {total_n_articles})'))) %>% 
  right_join(archy_cran_cites_all_areas_top_journals_by_year,
             by = "SO")

@@

all_archaeology_citing_R_over_time <- 
ggplot(archy_cran_cites_all_areas_top_journals_by_year,
       aes(PY,
           prop,
           colour = SO)) +
  geom_line(size = 2) +
  geom_point() +
  geom_text_repel(
    data = archy_cran_cites_all_areas_top_journals_by_year %>% 
      group_by(SO) %>% 
      slice(which.max(PY)) , 
    aes(label = SO_n),
    direction = "y",
    size = 5,
    nudge_x = 1,
    hjust = 0) +
  theme_minimal() +
  theme(legend.position="none",
        axis.text.x = element_text(angle = 90, 
                                   hjust = 1, 
                                   vjust = 0.5)) +
  scale_x_continuous(breaks = min_y:max_y,
                     labels = c(min_y:2017, 
                                rep("",
                                    length(2018:max_y))),
                     limits = c(min_y, max_y)) +
  scale_y_continuous(labels = scales::percent,
                     limits = c(0, 0.2)) +
  xlab("Publication year") +
  ylab("Percentage of articles in that journal")

ggsave(plot = all_archaeology_citing_R_over_time, 
       filename = "all_archaeology_citing_R_over_time.png",
       path = str_glue('{here::here()}/analysis/figures'),
       height = 7,
       width = 11)
```

```{r j-arch-sci-articles-citing-R}
# look at the trend in JAS, where there are the most articles

how_many_articles_in_jas <-  
  archy_cran_cites_all_areas_top_journals_by_year$total_n_articles[1]

jas_cites_r <- 
archy_cran_cites_all_areas_top_journals_by_year %>% 
  filter(SO == "JOURNAL OF ARCHAEOLOGICAL SCIENCE") %>% 
  ggplot(aes(PY,
           prop)) +
  geom_point(size = 4) +
  geom_smooth(method = "lm",
              se = FALSE) +
  stat_poly_eq(aes(label =  paste(..eq.label.., 
                                  ..adj.rr.label..,
                                  sep = "~~~~")),
               formula = y~x, 
               parse = TRUE,
               size = 3,
               label.y.npc = 0.7) +
  stat_fit_glance(label.y.npc = 0.7,
                  size = 3,
                  method = "lm", 
                  method.args = list(formula = y ~ x),
                  geom = "text",
                  aes(label = paste("p-value: ",
                                    signif(..p.value.., 
                                           digits = 4)))) +
  xlab("year") +
  ylab("Proportion of articles citing R") +
  ggtitle(expression(paste("Articles citing R in ", 
                           italic("Journal of Archaeological Science")))) +
  theme_bw(base_size = 6)
```

We conducted a similar bibliometric analysis for archaeology journals indexed by the Web of Science service. We used the 'Cited Reference Search' to find articles citing R in the same way as above, and then refined the results to keep only those articles published during 2007-2017 that are included in the Web of Science category 'Archaeology'. This resulted in `r how_many_articles_in_top_archy_journals` articles, of which `r total_number_archaeology_articles_citing_r` cite R.  Figure \@ref(fig:rchaeology-journals-with-subplot) shows the temporal trend of citations of R in archaeological articles. There are three interesting details revealed by this figure.

```{r archaeology-journals-with-subplot, fig.cap=str_glue('Percentage of Archaeology articles per year citing R ({total_number_archaeology_articles_citing_r} out of {prettyNum(how_many_articles_in_top_archy_journals, ",")} articles in sample. Data from Web of Science for 2007-2017')}

# combine main plot and subplot
all_archaeology_citing_R_over_time_and_subplot <- 
all_archaeology_citing_R_over_time  +
  annotation_custom(grob = ggplotGrob(jas_cites_r), 
                          xmin = 2030, xmax = 2040, 
                          ymin = 0.10, ymax = 0.17)

ggsave(plot = all_archaeology_citing_R_over_time_and_subplot, 
       filename = "all_archaeology_citing_R_over_time_and_subplot.png",
       path = str_glue('{here::here()}/analysis/figures'),
       height = 9,
       width = 12)

knitr::include_graphics(str_glue('{here::here()}/analysis/figures/all_archaeology_citing_R_over_time_and_subplot.png'))
```

The first detail we see here is that overall, the proportions and the absolute number of articles citing R in archaeology are much smaller than what we see in the ecology journals in Figure \@ref(fig:ecology-citing-r). Only the _Journal of Archaeological Science_ has more than 50 articles in our sample. We might conclude from this observation that archaeologists have yet to adopt programming for data analysis and visualisation in the same way that ecologists have.

The second detail is that most archaeology journals do not show any increase in the percentage of article citing R over time . The ecology journals shown in Figure \@ref(fig:ecology-citing-r) show clear upward trends of increasing proportions of articles citing R over time, but we do not see any clear trends in the archaeology articles considered together. Indeed, in several of the archaeological journals in our sample, the first citations of R only occured in the last two years. The sample sizes for the archaeology journals are too small to confidently infer any trend of over time, with the exception of the _Journal of Archaeological Science_ with `r how_many_articles_in_jas` articles. The inset plot in Figure \@ref(fig:rchaeology-journals-with-subplot) shows a statistical test of change over time for articles in the _Journal of Archaeological Science_. With a moderate r-squared value and low p-value for the linear model, we can conclude that there is some evidence for an increase in citations of R over time in this journal. This may indicate an increasing use of R among archaeologists, especially those working at the intersection of the natural and physical sciences. This may hint at the start of a trend similar to what we see in the ecology journals of a widespread

Following on from this observation about the increasing popularity of R in then _Journal of Archaeological Science_, the third detail we see in these results is the distinctive types of journals that have articles citing R. There is a clear focus on journals that publich scientific archaeology, especially those that focus on specialised empirical analysis. In addition to the general archaeology journals, we also see _Lithic Technology_, _Anthropozoologica_, and _International Journal of Osteology_ in the journal names, indicating that we find R used by specialists in stone artefacts and faunal remains. To further investigate the topics that archaeologists are useing R to study, we conducted a statistical analysis of the words in the titles of all the articles in our sample.

## What are R-using archaeologists writing about?

```{r text-mining-article-titles-archaeology}

# https://www.tidytextmining.com/twitter.html

@@

wc_archaeology_2007_to_2017_no_R_cites <- 
  archy_all_years %>% 
  filter(TI %nin% archy_cran_all_years$TI)

# combine
archy_journal_articles <- 
  bind_rows(wc_archaeology_2007_to_2017_no_R_cites %>% 
            mutate(ID = "no_R_cites",
                   VL = as.character(VL)),
           archy_cran_all_years %>% 
            mutate(ID = "has_R_cites"))

tidy_cites <- 
archy_journal_articles %>% 
  unnest_tokens(word, TI) %>%
  filter(!word %in% c(stop_words$word, "archaeology", 
                      "archaeological",
                      "analysis", "study", "site", "period", "past", "patterns", "investigation", "analyses", "late", "region", "based", "central", "north", "south", "east", "west", "western", "eastern", "ad", "pre", "sites", "evidence", "prehistoric", "remains", "change", "de", "identification", "northern", "southern", "valley", "implications", "valley", "history", "historical", "modern", "context", "stable", "middle", "approach", "san", "artifacts", "palaeolithic"),
         str_detect(word, "[a-z]"))

@@

Figure \@ref(fig:text-mining-article-titles-archaeology-plot) supports our initial observations that archaeologists using R tend to be doing archaeology that involves the natural and physical sciences. Articles that cite R more often have titles that include terms such as 'lithic', 'experimental', and 'isotope'. On the other hand, articles that do not cite R more often discuss the Roman and Bronze Age periods, and less often reflect a technical specialisation, if there is one, it seems to ceramics.

## From using code in research to sharing code with publications

```{r archaeology-reproducible-articles}

archy_ctv_readme <- readLines(str_glue('{here::here()}/analysis/data/raw_data/ctv-archaeology-README.md'))
